\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-17 chapters/covert_align.tex
Proofread mathematical notation
Indicated comments Reinhold and Fabien

\chapter{Compensating \acs{erp} jitter for gaze-independence}%
\label{sec:covert-align}
\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}

@@

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert \ac{vsa} compared to overt
\ac{vsa}.
This latency variability contributes to reduced covert \ac{vsa} decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert \ac{vsa}
decoding, but did not verify this hypothesis directly.
Additionally, \textcite{Hardiansyah2020} developed a classifier for
covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial \ac{erp} component latency
features on covert \ac{vsa} inference, yet did not attempt to correct the amplitude
features for these latencies.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_3.tikz.tex}
    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
    settings, the
    participant is cued to fixate on one of the targets.}%
    \label{fig:covert-align/interface/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_4.tikz.tex}
    \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
    cued to fixate at the center of the interface.}%
    \label{fig:covert-align/interface/covert}
  \end{subfigure}
\end{figure}

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in thdeep q learning tutoriale BNCI2014-009 dataset
for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.

Overall, we observed a 5.10\%pt. accuracy increase with \ac{wcble} over
\ac{tlda} for covert \ac{vsa} in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
These results compare to the performance gain in \cite{Zisk2022}.
They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
step-wise Linear Discriminant Analysis  (SWLDA) for 6 participants with
\ac{als}, whose SWLDA performance also suffered from jitter.
Note that interpretation of this comparison may be challenging due to differences in
interface design (number of targets, \ac{isi}), subject population,
\ac{eeg} recording procedure, and available training data.

@@

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon rank-sum tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions.
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding.

@@

We believe the relative increase in performance of our proposed decoder in covert and
split \ac{vsa}, and the lack thereof in overt \ac{vsa}, could stem from the following:
\begin{enumerate*}[label=(\arabic*)]
	\item Covert and split \ac{vsa} exhibit higher P3 jitter than overt \ac{vsa}.
        In covert and split \ac{vsa}, participants have to execute a dual task by
        dissociating their visuospatial attention and gaze fixation.
        Evidence shows that \ac{erp} latency variability is higher when attention is
        divided~\cite{Polich2007,Arico2014}.
        \textcite{Arico2014} also partly attribute higher latency jitter to the covert \ac{vsa}
        task performed in the BNCI2014-009 dataset, since the GeoSpell interface
        requires both spatial and feature attention.
  \item In overt \ac{vsa}, the first-stage classifier can rely mostly on the modulation
        of early visually evoked potentials (VEPs) like N1 rather than on the
        P3~\cite{Treder2010}.
        These VEPs are closely related to visual processing, hence exhibit
        lower jitter, contrary to P3, which is more prone to the effects of
        attention and workload~\cite{Hu2010}, reducing the contribution of
        alignment.
  \item This property can also result in the estimation of VEP latencies
        instead of the P3 latency, and \ac{wcble} would in this case fail to
        increase the P3 \ac{snr}, which still could be somewhat jittered in overt
        \ac{vsa}.
  \item Aligning to the P3 will lower the \ac{snr} of early VEPs, while aligning to
        VEPs will lower P3 \ac{snr}, since they are not time-locked to each other.
	\item Covert and split \ac{vsa} \acp{erp} may exhibit lower \ac{snr} than overt
        \ac{vsa} due to lower P3 amplitudes or even due to the presence of higher P3
        jitter itself.
        Higher \ac{snr} in overt \ac{vsa} results in higher decoding performance of
        state-of-the-art classifiers, leaving less room for relative improvement
        in this case.
\end{enumerate*}

Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
present in covert and split \ac{vsa} compared to overt \ac{vsa},
we justify our approach similarly to \cite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes, while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points toward higher P3 jitter as the main
contributing factor.
While \autoref{fig:erps} shows no visible smearing effect in
the shape of the \acp{erp}, a more quantitative analysis on latencies in
\autoref{fig:jitter} indicates the opposite.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
range (IQR) of the estimated latencies of target trials, revealed that overt
\ac{vsa} exhibited significantly lower jitter than all other conditions for both datasets.
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally, \textcite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert \ac{vsa} by comparing performance between
overt and covert \ac{vsa}, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert \ac{vsa} decoding performance.
However, the large increase in performance for covert \ac{vsa} in \cite{Arico2014} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.

>>>>

\chapter{Compensating \acs{erp} jitter for gaze-independence}%
\label{sec:covert-align}
\epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ A
Scandal in Bohemia -- Sir Arthur Conan Doyle}
\noindent\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}

@@

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert \ac{vsa} compared to overt
\ac{vsa}\todo{qualitative, discuss}.
This latency variability contributes to reduced covert \ac{vsa} decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert \ac{vsa}
decoding, but did not verify this hypothesis directly.
Additionally, \textcite{Hardiansyah2020} developed a classifier for
covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial \ac{erp} component latency
features on covert \ac{vsa} inference, yet did not attempt to correct the amplitude
features for these latencies.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_3.tikz.tex}
    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
    settings, the
    participant is cued to fixate on one of the targets.}%
    \label{fig:covert-align/interface/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_4.tikz.tex}
    \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
    cued to fixate at the center of the interface.}%
    \label{fig:covert-align/interface/covert}
  \end{subfigure}
  \caption[Stimulation interface layout.]{Layout and visual elements of the experimental stimulation
  interface used in recording the CVSA-ERP dataset.}
\end{figure}

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.\todo{qualitative}
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.\todo{qualitative}
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.\todo{qualitative}

Overall, we observed a 5.10\%pt. accuracy increase with \ac{wcble} over
\ac{tlda} for covert \ac{vsa} in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
These results compare to the performance gain reported by \textcite{Zisk2022}.
They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
step-wise Linear Discriminant Analysis  (SWLDA) for 6 participants with
\ac{als}, whose SWLDA performance also suffered from jitter.
Note that interpretation of this comparison may be challenging due to differences in
interface design (number of targets, \ac{isi}), subject population,
\ac{eeg} recording procedure, and available training data.

@@

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon rank-sum tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions.\todo{qualitative}
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding.\todo{qualitative}

@@

We believe the relative increase in performance of our proposed decoder in covert and
split \ac{vsa}, and the lack thereof in overt \ac{vsa}, could stem from the following:
\begin{enumerate*}[label=(\arabic*)]
  \item Covert and split \ac{vsa} exhibit higher P3 jitter than overt \ac{vsa}.
        In covert and split \ac{vsa}, participants have to execute a dual task by
        dissociating their visuospatial attention and gaze fixation.
        Evidence shows that \ac{erp} latency variability is
        higher when attention is
        divided~\cite{Polich2007,Arico2014}.
        \textcite{Arico2014} also partly attribute higher latency jitter to the covert \ac{vsa}
        task performed in the BNCI2014-009 dataset, since the GeoSpell interface
        requires both spatial and feature attention.
  \item In overt \ac{vsa}, the first-stage classifier can rely mostly on the modulation
        of early visually evoked potentials (VEPs) like N1 rather than on the
        P3~\cite{Treder2010}.
        These VEPs are closely related to visual processing, hence exhibit
        lower jitter, contrary to P3, which is more prone to the effects of
        attention and workload~\cite{Hu2010}, reducing the contribution of
        alignment.
  \item This property can also result in the estimation of VEP latencies
        instead of the P3 latency, and \ac{wcble} would in this case fail to
        increase the P3 \ac{snr}, which still could be somewhat jittered in overt
        \ac{vsa}.
  \item Aligning to the P3 will lower the \ac{snr} of early VEPs, while aligning to
        VEPs will lower P3 \ac{snr}, since they are not time-locked to each other.
	\item Covert and split \ac{vsa} \acp{erp} may exhibit lower \ac{snr} than overt
        \ac{vsa} due to lower P3 amplitudes or even due to the presence of higher P3
        jitter itself.
        Higher \ac{snr} in overt \ac{vsa} results in higher decoding performance of
        state-of-the-art classifiers, leaving less room for relative improvement
        in this case.
\end{enumerate*}
\todo{first item qualitative}

Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
present in covert and split \ac{vsa} compared to overt \ac{vsa},
we justify our approach similarly to \cite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes, while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points toward higher P3 jitter as the main
contributing factor.
While \autoref{fig:erps} shows no visible smearing effect in
the shape of the \acp{erp}, a more quantitative analysis on latencies in
\autoref{fig:jitter} indicates the opposite.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
range (IQR) of the estimated latencies of target trials, revealed that overt
\ac{vsa} exhibited significantly lower jitter than all other conditions for
both datasets.\todo{qualitative}
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally, \textcite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert \ac{vsa} by comparing performance between
overt and covert \ac{vsa}, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert \ac{vsa} decoding performance.
However, the large increase in performance for covert \ac{vsa} reported by
\cite{Arico2014}\todo{qualitative} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.\todo{qualitative}

### 2024-11-17 chapters/gaze_independence.tex
Indicated comments Reinhold and Fabien

Some other groups that are heavily affected are \ac{cp}~\cite{Fazzi2012} and
\ac{fa}~\cite{Fahey2008,Hocking2010,Furman1983}.
Individuals with some neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

@@

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection but also result in lower information transfer
rates compared to visual paradigms.
While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

This leads us to the second approach: current visual stimulation paradigms can be optimized so that
the stimuli are always present in the field of view, either overtly~\cite{Acqualagna2013, Won2018, Lin2018} or covertly~\cite{Pires2011,Lees2018}.
Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm, in particular, is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the user's field of view.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

@@

More recent work has made advances in decoding lateralized covert
\ac{vsa} by harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}.
The amplitude of the N2pc component in the contralateral hemisphere is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56 bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used a few stimulation targets.
In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance may be to increase overall \ac{erp}
decoding performance.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
      The user is free to direct their gaze as they deem most comfortable.
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Acl{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze/vsa}
\end{figure}

>>>>

Another other group that is heavily affected, is \ac{cp}~\cite{Fazzi2012}.
Additionally, individuals with other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

@@

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection but also result in lower \acp{itr}
compared to visual paradigms\todo{qualitative}.
While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent
variability~\cite{Severens2014,Reichert2020b}.\todo{qualitative}
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

This leads us to the second approach: current visual stimulation paradigms can be optimized so that
the stimuli are always present in the field of view, either overtly~\cite{Acqualagna2013, Won2018, Lin2018} or covertly~\cite{Pires2011,Lees2018}.
Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm, in particular, is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the user's field of view.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small
sets of targets (e.g. 3) are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

@@

More recent work has made advances in decoding lateralized covert
\ac{vsa} by harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}.
The amplitude of the N2pc component in the contralateral hemisphere is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56 bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used a few stimulation targets.\todo{qualitative}
In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings\todo{qualitative}.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance may be to increase overall \ac{erp}
decoding performance.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      The user is free to direct their gaze as they deem most comfortable.
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Acl{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze/vsa}
\end{figure}
